---
title: "Victoria Line Train Service Optimisation"
author: "Zino Harrison"
date: "2026-09-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(ggplot2)
library(tidyverse)
library(patchwork)
library(kableExtra)
```

## 1. Introduction

The aim of this project is to determine the optimal train allocation across services on the Victoria Line of the London Underground to minimise passenger overcrowding, using TfL NUMBAT data. An integer linear programming model is developed to find the minimum number of trains that can be dispatched in each interval and direction whilst maintaining zero overcrowding. 

The initial model however only minimised overcrowding. A sensitivity analysis, investigating the effect of a substantial reduction in train capacity revealed a solution where overcrowding was zero could still be found. Therefore the model was revised to minimise train dispatches and a sensitivity analysis was performed, investigating the effect of a decrease in train capacity on the minimum number of trains that can be dispatched, whilst maintaining overcrowding at zero.

The model assumes a fixed number of cars per train and average capacity per car. Trains on the Victoria Line consist of 8 cars with an average capacity per car of 107, 856 is assumed to be the average capacity per train. The Victoria Line fleet size is assumed to be 47 trains and is incorporated as a fleet constraint in the model.

## 2. Data

### 2.1 Data source

The model uses data from TfL NUMBAT 2025 
\newline
<https://crowding.data.tfl.gov.uk/>.
\newline
The data on a typical autumn monday (NBT25MON) was chosen and Link Loads data was used.

Information used in determining the model constraints was gathered from the following websites:
\newline
1. <https://tfl.gov.uk/corporate/transparency/freedom-of-information/foi-request-detail?referenceId=FOI-4157-2425>
\newline
2. <https://content.tfl.gov.uk/06-lu-rolling-stock.pdf>
\newline
3. <https://mycommute.london/victoria>

### 2.2 Data preparation

In the data preparation phase columns not relevant to the model were dropped and the Victoria Line was selected. Then the data frame was converted into a long format, time intervals were initially arranged in columns but the process converted them to rows. Finally the order link numbers were then adjusted - they were kept the same for NB but reversed for SB as instead of order the numbers were going to be allocated to links. 

\newpage

## 3. Mathematical Model

## 3.1 Assumptions

The NUMBAT data used 15 minute intervals, which significantly influenced the design of the model. The travel time between each station is roughly 2 minutes and the total time it takes to reach one end of the line to the other is approximately 31 minutes. However for simplicity this was assumed to be 30 minutes (and therfore 2 intervals). 

The model modeled trains as completing a cycle around the line (starting from one end, reaching the end, turning around and then reaching the other end). However turnaround time was not taken into consideration in the model and it was instead assumed it would take 60 minutes (4 intervals) for a train to complete a full cycle.

To remain consistent with the 15 minute intervals, in each direction the line was split into halves from station. Four trains are assumed to be in circulation at t = 0, one in each half. This was to ensure overcrowding remained at zero at all times.

Additionally it is assumed all trains are identical with a fixed capacity of 856.

Link load was used to determine passenger demand and by setting overcrowding equal to zero, all passenger demand must be accommodated for in each interval.

Finally it is assumed there are no delays or cancellations.

## 3.2 Model 1
$$
\text{Let } t = 1,...,T \text{ denote each 15-minute interval}
$$
$$
\text{Let} I = {1,...,15} \text{ where } i \in I \text{ denotes the link number}
$$
	
	1. Brixton - Stockwell
	2. Stockwell - Vauxhall
	3. Vauxhall - Pimlico
	4. Pimlico - Victoria
	5. Victoria - Green Park
	6. Green Park - Oxford Circus
	7. Oxford Circus - Warren Street
	8. Warren Street - Euston
	9. Euston - King’s Cross St Pancras 
	10. King’s Cross St Pancras  - Highbury & Islington
	11. Highbury & Islington - Finsbury Park
	12. Finsbury Park - Seven Sisters
	13. Seven Sisters - Tottenham Hale
	14. Tottenham Hale - Blackhorse Road
	15. Blackhorse Road - Walthamstow Central 

### Decision variables

$$
O_{i,t}^{d} - \text{ overcrowding in direction } d \in \{\mathrm{NB}, \mathrm{SB}\} \text{ at link } i \text{ during interval } t
$$
$$
x_{t}^{NB} - \text{the number of trains dispatched from Brixton during interval } t
$$
$$
x_{t}^{SB} - \text{the number of trains dispatched from Walthamstow Central during interval } t
$$

### Parameters

$$
L_{i,t}^d - \text{the number of passengers travelling on link } i \text{ in direction } d \in \{\mathrm{NB}, \mathrm{SB}\}
$$
$$
c - \text{capacity per train}
$$
$$
F - \text{fleet size}
$$

### Objective function

$$
\text{Minimise } 
\sum_{d \in \{\mathrm{NB}, \mathrm{SB}\}}
\sum_{i=1}^{15} 
\sum_t^T O_{i,t}^d
$$

### Train circulation

Trains are modeled following a cyclical structure. Each train is either dispatched at Brixton or Walthamstow Central and follows the route until the end of the line and then turns around to reach the other end of the line, thus completing a cycle. It is assumed it takes 30 minutes to travel the full line, so takes 60 minutes to complete a cycle.

The line is split into halves to be compatible with the 15-minute intervals used in the NUMBAT dataset.

It is also assumes at t = 0 there is one train in circulation in each half of the line in each direction.

$$
x_t^{NB} = x_t^{SB} = 0  \text{ for } t \leq -2
$$
$$
x_t^{NB} = x_t^{SB} = 1 \text{ for } t \in \{\mathrm{-1}, \mathrm{0}\}
$$

### Capacity constraints

Link load is the demand on trains between platforms. The model defines overcrowding as the excess passenger demand over the provided capacity in the relevant link and interval.

$$
O_{i,t}^{NB} 
\geq
L_{i,t}^{NB} - c(x_t^{NB} + x_{t-2}^{SB})
\text{ }, \forall (i \leq 8 \cap \text{ } t >2)
$$

$$
O_{i,t}^{NB}
\geq
L_{i,t}^{NB} - c(x_{t-1}^{NB} + x_{t-3}^{SB})
\text{ }, \forall (i \geq 9 \cap \text{ } t >3)
$$

$$
O_{i,t}^{SB} 
\geq
L_{i,t}^{SB} - c(x_{t}^{SB} + x_{t-2}^{NB})
\text{ }, \forall (i \geq 9 \cap \text{ } t >2)
$$

$$
O_{i,t}^{SB} 
\geq
L_{i,t}^{SB} - c(x_{t-1}^{SB} + x_{t-3}^{NB})
\text{ }, \forall (i \leq 8 \cap  \text{ } t >3)
$$

### Fleet constraint

$$
\sum_{\tau = t-3}^t 
( x_{\tau}^{NB} + x_{\tau}^{SB} )
\leq 
F, 
\text{ } 
\forall t
$$

### Frequency constraint

Peak frequency (the most trains on the newtork - in one direction) is 36 trains per hour. Since the model used 15 minute intervals we have $$ 36 \times \frac{15}{60} = 9 $$

$$
x_t^{NB} \leq 9 \text{ and } x_t^{SB} \leq 9
$$

### Integer constrain

$$
x_t^d \in \mathbb{Z}_{\geq 0}
$$

## 3.3 Model 2

The second model retains the train circulation, fleet and frequency constraints from model 1. However it removes the overcrowding constraint, essentially setting it equal to zero so it requires  passenger demand to be no greater than available capacity.

### Decision variables


$$
x_{t}^{NB} - \text{the number of trains dispatched from Brixton during interval } t
$$
$$
x_{t}^{SB} - \text{the number of trains dispatched from Walthamstow Central during interval } t
$$

### Parameters

$$
L_{i,t}^d - \text{the number of passengers travelling on link } i \text{ in direction } d \in \{\mathrm{NB}, \mathrm{SB}\}
$$
$$
c - \text{capacity per train}
$$
$$
F - \text{fleet size}
$$

### Objective function

$$
\text{Minimise } 
\sum_{d \in \{\mathrm{NB}, \mathrm{SB}\}}
\sum_{t=1}^T x_{t}^d
$$

### Constrains

$$
x_t^{NB} = x_t^{SB} = 0  \text{ for } t \leq -2
$$
$$
x_t^{NB} = x_t^{SB} = 1 \text{ for } t \in \{\mathrm{-1}, \mathrm{0}\}
$$
$$
L_{i,t}^{NB} \leq c(x_t^{NB} + x_{t-2}^{SB})
\text{ }, \forall (i \leq 8 \cap \text{ } t >2)
$$
$$
L_{i,t}^{NB} \leq c(x_{t-1}^{NB} + x_{t-3}^{SB})
\text{ }, \forall (i \geq 9 \cap \text{ } t >3)
$$
$$
L_{i,t}^{SB} \leq c(x_{t}^{SB} + x_{t-2}^{NB})
\text{ }, \forall (i \geq 9 \cap \text{ } t >2)
$$
$$
L_{i,t}^{SB} \leq c(x_{t-1}^{SB} + x_{t-3}^{NB})
\text{ }, \forall (i \leq 8 \cap  \text{ } t >3)
$$
$$
4 + \sum_{\tau = t-3}^t 
( x_{\tau}^{NB} + x_{\tau}^{SB} )
\leq 
F, 
\text{ } 
\forall t
$$
$$
x_t^{NB} \leq 9
$$
$$
x_t^{SB} \leq 9
$$
$$
x_t^d \in \mathbb{Z}_{\geq 0}
$$

## 4. Optimisation Approach

The model was implemented in Python using Pyomo and solved using GLPK for the first model. However the solution time became impractical for the second model so HiGHS was used instead. The initial model was implemented and a solution where overcrowding equaled zero was found for the baseline capacity of 856. However after adjusting the capacity down to even 428, the model still found a solution where overcrowding was equal to zero. Therefore the model was instead adjusted to minimise the trains dispatches needed for overcrowding to remain at zero. 

\newpage

## 5. Results

### 5.1 Optimal dispatch schedule and Active fleet

```{r, echo=FALSE}
schedule <- read_csv("~/Tube service optimisation/data/processed/Optimised_m2_results.csv", 
                     show_col_types = FALSE)

kable(head(schedule, 15),
      caption = "First 15 intervals of the optimised train dispatch schedule at capacity 856")
```


```{r, echo=FALSE}
vic_mon <- read_csv("~/Tube service optimisation/data/processed/Optimised_m2_results.csv", 
                     show_col_types = FALSE)


# Combine directions into one column

vic_mon_merged <- pivot_longer(vic_mon, 
                               cols = c('NB', 'SB'),
                               names_to = 'Direction',
                               values_to = 'Trains_Dispatched')

# Fix times

vic_mon_merged$Start_Time <- vic_mon_merged$Time 

vic_mon_merged <- vic_mon_merged %>% 
  separate(Start_Time, into = c("Start_Time", "End Time"), sep = "-")

vic_mon_merged$Start_Time <- hms::as_hms(
  paste0(substr(vic_mon_merged$Start_Time, 1, 2),":",
    substr(vic_mon_merged$Start_Time, 3, 4),":00"
  )
)



#df for direction
dispatch <- vic_mon_merged[c("Start_Time", "Direction", 
                                  "Trains_Dispatched", "t")]

# df for active
active <- vic_mon_merged[c("Start_Time", "Trains_active", 
                                    "Trains_Dispatched", "t")]


#Plot

ggplot(dispatch, aes(x = t, y = Trains_Dispatched, 
                           group = Direction, 
                     colour = Direction)) + 
  geom_step(linewidth = 1) + 
  geom_point(size = 1) + 
  geom_step(data = active, aes(x = t,
                               y = Trains_active,
                               colour = "Active Trains"),
            inherit.aes = FALSE,
            linewidth = 1
    ) + geom_point(data = active, size = 1, inherit.aes = FALSE,
                   aes(x = t,
                               y = Trains_active,
                               colour = "Active Trains")) + 
  scale_x_continuous(breaks = active$t[seq(1, nrow(active), by = 20)],
    labels = active$Start_Time[seq(1, nrow(active), by = 20)]
  ) +
  theme_minimal() + 
  labs(
    title = "Optimised Train Dispatches",
    x = "15 minute interval",
    y = "Number of Trains") +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "top")


```
The plot shows the optimised number of trains in circulation in each 15 minute interval (assuming capacity is 856). The number of trains being dispatched fluctuates, although the number of trains in circulation increases to meet peak demand in the morning. The number of active trains peaks around 08:00/08:30 at 23 in circulation. After the initial rush, link load falls so the number of trains being dispatched during each interval falls and so does the number of active trains decreasing. The  number of active trains continues to fall until just after 12:30, where it remains at 8 trains active on he line until just before 15:00. From this point the number of trains dispatched in both directions increases and the number of trains in circulation increases and peaks at 24 at around 17:00, and begins to fall again at around 19:00. The numberd of active trains continues to fall until just after 20:00, when it then plateaus at 8 in circulation, then falling again after 22:30, eventually falling to 0 after 01:00 in the morning. It can be seen that more trains are dispatched on the Southbound line, especially during peak times. Additionally it it only after 20:00 that during some intervals no trains are dispatched on the Southbound line (with the exception of the first two interval after 05:00). However on the Northbound line between the rough hours of 11:00 and 14:30 no trains are dispatched.


## 6. Sensitivity Analysis

### 6.1 Effect of train capacity

```{r, echo=FALSE}
## Bar plot of total trains dispatched vs train capacity

min_total <-  read_csv("~/Tube service optimisation/data/processed/Trains_Sensitivity.csv", 
                       show_col_types = FALSE)

min_total$Capacity <- as.character(min_total$Capacity)

ggplot(min_total, aes(x = Capacity, y = Minimised_trains,
                      fill = Capacity,
                      colour = Capacity)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Trains dispatched vs Train Capacity",
       x = "Train Capacity", y = "Total Trains Dispatched") +
  theme_minimal()
```
The plot suggests as capacity falls, the number of trains dispatched must rise in order to meet the constraint of zero overcrowding. At capacity 856, used in the model, the total number of trains dispatched was just below 250. But when capacity is reduced by 50% to 428 the total number of trains dispatched is just below 450 (an increase of around 200). Each time capacity is decreased by 12% the increase in the total number of dispatched trains rises slightly more. For example the increase in train dispatches when capacity decreases from 535 to 428 in much larger than the increase in trains when capacity is decreased from 856 to 749.

### 6.2 Effect on fleet utilisation

```{r, echo=FALSE, fig.width=7.5, fig.height=10}

vic_mon <- read_csv("~/Tube service optimisation/data/processed/Sensitivity_m2.csv",
                    show_col_types = FALSE)

# Direction to one column

vic_mon <- pivot_longer(vic_mon, 
                               cols = c('NB', 'SB'),
                               names_to = 'Direction',
                               values_to = 'Trains_Dispatched')

# Fix times

vic_mon$Start_Time <- vic_mon$Time 

vic_mon <- vic_mon %>% 
  separate(Start_Time, into = c("Start_Time", "End Time"), sep = "-")

vic_mon$Start_Time <- paste0(
  substr(vic_mon$Start_Time, 1, 2),
  ":",
  substr(vic_mon$Start_Time, 3, 4)
)

vic_mon$Capacity <- as.character(vic_mon$Capacity)

# Dispatch schedule for selected capacities

#df for direction
dispatch <- vic_mon[c("Start_Time", "Direction", 
                             "Trains_Dispatched", "t", "Capacity")]

# df for active
active <- vic_mon[c("Start_Time", "Trains_active", 
                           "Trains_Dispatched", "t", "Capacity")]

# create function for plot

plot_capacity <- function(capacity) {
  
  plot_data1 <- dispatch %>%
    filter(Capacity == capacity)
  
  plot_data2 <- active %>%
    filter(Capacity == capacity)
  
  ggplot(plot_data1, aes(x = t, y = Trains_Dispatched, 
                       group = Direction, 
                       colour = Direction)) + 
    geom_step(linewidth = 0.5) + 
    geom_point(size = 0.5) + 
    geom_step(data = plot_data2, aes(x = t,
                                 y = Trains_active,
                                 colour = "Active Trains"),
              inherit.aes = FALSE,
              linewidth = 1
    ) + geom_point(data = plot_data2, size = 1, inherit.aes = FALSE,
                   aes(x = t,
                       y = Trains_active,
                       colour = "Active Trains")) + 
    scale_x_continuous(breaks = plot_data2$t[seq(1, nrow(plot_data2), by = 20)],
                       labels = plot_data2$Start_Time[seq(1, nrow(plot_data2), by = 20)]
    ) +
    scale_y_continuous(limits = c(0, 50), breaks = seq(0,50, 10)) +
    theme_minimal() + 
    labs(
      title = paste("Optimised Train Dispatches at capacity", capacity),
      x = "15 minute interval",
      y = "Number of Trains") +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "top")
}

p1 <- plot_capacity(856)
p2 <- plot_capacity(749)
p3 <- plot_capacity(642)
p4 <- plot_capacity(428)

combined1 <- p1 / p2

combined2 <- p3 / p4

```

```{r, echo=FALSE, fig.width=7.5, fig.height=10}
combined1
```

```{r, echo=FALSE, fig.width=7.5, fig.height=10}
combined2
```

```{r, echo=FALSE}

# Plot trains active for each interval, split by capacity

ggplot(vic_mon, aes(x = t, 
                    y = Trains_active,
                    group = Capacity, 
                    colour = Capacity)) + 
  geom_step(linewidth = 1) + 
  geom_point(size = 1) + 
  labs(
    title = "Minimised Active Trains for each Capacity and Interval",
    x = "Time",
    y = "Number of Trains Active"
  ) +
  scale_x_continuous(breaks = vic_mon$t[seq(1, nrow(vic_mon), by = 80)],
                     labels = vic_mon$Start_Time[seq(1, nrow(vic_mon), by = 80)]) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "top"
  )
```

The first four plots show the number of active trains and trains dispatched in each direction across the capacities 856, 749, 642, 428. Meanwhile the plot below shows the number of active trains in each interval for the chosen capacities on one plot, providing a more direct comparison.

The plots show a steeper increase in active trains during peak times for the lower capacities. The individual plots also highlight how for the lower capacities a greater number of trains need to be dispatched in order to meet the increased passenger demand. For example at capacity 428 there are several intervals where 9 trains are dispatched in one direction (primarily SB) whilst the maximum number of trains dispatched in one direction during one interval for capacity 856 is 6. Even betweeen peak times, the number of active trains is greater at lower capacities (at capacity 428 it falls to 16 but at capacity 856 it falls to 8).

However across the different capacities, the same pattern of the number of active trains peaking during peak times in the morning and evening can be observed, since they are all responding to the same link load data. The key difference is that at lower capacities a greater number of trains must be dispatched in order to meet this demand.

## 7. Limitations

## 8. Conclusion
