---
title: "Victoria Line Train Service Optimisation"
author: "Zino Harrison"
date: "2026-09-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(ggplot2)
library(tidyverse)
library(patchwork)
library(kableExtra)
```

## 1. Introduction

The aim of this project is to determine the optimal train allocation across services on the Victoria Line of the London Underground to minimise passenger overcrowding, using TfL NUMBAT data. An integer linear programming model is developed to find the minimum number of trains that can be dispatched in each interval and direction whilst maintaining zero overcrowding. 

The initial model however only minimised overcrowding. A sensitivity analysis, investigbating the effect of a substantial reduction in train capacity revealed a solution where overcrowding was zero could still be found. Therefore the model was revised to minimise train dispatched instead and a sensitivity analysis was performed, investigationg the effect of a decrease in train capacity on the minimum number of trains that can be dispatched so that overcrowding is zero.

The model assumes a fixed number of cars per train and average capacity per car. Trains on the Victoria Line consist of 8 cars with an average capacity per car of 107, so the average capacity per train is calculated to be 856. Additionally the victoria Line fleet size is 47, which is used as a constraint in the model.f

## 2. Data

### 2.1 Data source

The model uses data from TfL NUMBAT 2025 
\newline
<https://crowding.data.tfl.gov.uk/>.
\newline
The data on a typical autumn monday (NBT25MON) was chosen, simply due to it being the first day in the week. Link Loads data was used.

Information used in determining the model constraints was gathered from the following websites:
\newline
1. <https://tfl.gov.uk/corporate/transparency/freedom-of-information/foi-request-detail?referenceId=FOI-4157-2425>
\newline
2. <https://content.tfl.gov.uk/06-lu-rolling-stock.pdf>
\newline
3. <https://mycommute.london/victoria>

### 2.2 Data preparation

In the data preparation phase columns not relevant to the model were dropped and the Victoria Line was selected. Then the data frame was converted to a long format, time intervals were initially arranged in columns but the process converted them to rows. Finally the order link numbers were then adjusted - they were kept the same for NB but reversed for SB as instead of order the numbers were going to be allocated to links. 


## 3. Mathematical Model

## 3.1 Model 1
$$
\text{Let } t = 1,...,T \text{ denote each 15-minute interval}
$$
$$
\text{Let} I = {1,...,15} \text{ where } i \in I \text{ denotes the link number}
$$
	
	1. Brixton - Stockwell
	2. Stockwell - Vauxhall
	3. Vauxhall - Pimlico
	4. Pimlico - Victoria
	5. Victoria - Green Park
	6. Green Park - Oxford Circus
	7. Oxford Circus - Warren Street
	8. Warren Street - Euston
	9. Euston - King’s Cross St Pancras 
	10. King’s Cross St Pancras  - Highbury & Islington
	11. Highbury & Islington - Finsbury Park
	12. Finsbury Park - Seven Sisters
	13. Seven Sisters - Tottenham Hale
	14. Tottenham Hale - Blackhorse Road
	15. Blackhorse Road - Walthamstow Central 

### Decision variables

$$
O_{i,t}^{d} - \text{ overcrowding in direction } d \in \{\mathrm{NB}, \mathrm{SB}\} \text{ at link } i \text{ during interval } t
$$
$$
x_{t}^{NB} - \text{the number of trains dispatched from Brixton during interval } t
$$
$$
x_{t}^{SB} - \text{the number of trains dispatched from Walthamstow Central during interval } t
$$

### Parameters

$$
L_{i,t}^d - \text{the number of passengers travelling on link } i \text{ in direction } d \in \{\mathrm{NB}, \mathrm{SB}\}
$$
$$
c - \text{capacity per train}
$$
$$
F - \text{fleet size}
$$

### Objective function

$$
\text{Minimise } 
\sum_{d \in \text{{NB,SB}}} 
\sum_{i=1}^{15} 
\sum_t^T O_{i,t}^d
$$

### Train circulation

Trains are modeled following a cyclical structure. Each train is either dispatched at Brixton or Walthamstow Central and follows the route until the end of the line and then turns around to reach the other end of the line, thus completing a cycle. It is assumed it takes 30 minutes to travel the full line, so takes 60 minutes to complete a cycle.

The line is split into halves to be compatible with the 15-minute intervals used in the NUMBAT dataset.

It is also assumes at t = 0 there is one train in circulation in each half of the line in each direction.

$$
x_t = x_t = 0  \text{ for } t \leq -2
$$
$$
x_t = x_t = 1 \text{ for } t \in \{\mathrm{-1}, \mathrm{0}\}
$$

### Capacity constraints

Link load is the demand on trains between platforms. The model defines overcrowding as the excess passenger demand over the provided capacity in the relevant link and interval.

$$
O_{i,t}^{NB} 
\geq
L_{i,t}^{NB} - c(x_t^{NB} + x_{t-2}^{SB})
\text{ }, \forall (i \leq 8 \cap \text{ } t >2)
$$

$$
O_{i,t}^{NB}
\geq
L_{i,t}^{NB} - c(x_{t-1}^{NB} + x_{t-3}^{SB})
\text{ }, \forall (i \geq 9 \cap \text{ } t >3)
$$

$$
O_{i,t}^{SB} 
\geq
L_{i,t}^{SB} - c(x_{t}^{SB} + x_{t-2}^{NB})
\text{ }, \forall (i \geq 9 \cap \text{ } t >2)
$$

$$
O_{i,t}^{SB} 
\geq
L_{i,t}^{SB} - c(x_{t-1}^{SB} + x_{t-3}^{NB})
\text{ }, \forall (i \leq 8 \cap  \text{ } t >3)
$$

### Fleet constraint

$$
4 + \sum_{\tau = t-3}^t 
( x_{\tau}^{NB} + x_{\tau}^{SB} )
\leq 
F, 
\text{ } 
\forall t
$$

### Frequency constraint

Peak frequency (the most trains on the newtork - in one direction) is 36 trains per hour. Since the model used 15 minute intervals we have $$ 36 \times \frac{15}{60} = 9 $$

$$
x_t^{NB} \leq 9 \text{ and } x_t^{SB} \leq 9
$$

### Integer constrain

$$
x_t^d \in \mathbb{Z}_{\geq 0}
$$

## 3.2 Model 2

The second model is very similar, following the same constraints with the addition of overcrowding being set to 0.

### Decision variables


$$
x_{t}^{NB} - \text{the number of trains dispatched from Brixton during interval } t
$$
$$
x_{t}^{SB} - \text{the number of trains dispatched from Walthamstow Central during interval } t
$$

### Parameters

$$
L_{i,t}^d - \text{the number of passengers travelling on link } i \text{ in direction } d \in \{\mathrm{NB}, \mathrm{SB}\}
$$
$$
c - \text{capacity per train}
$$
$$
F - \text{fleet size}
$$

### Objective function

$$
\text{Minimise } 
\sum_{d \in \text{{NB,SB}}} 
\sum_{t=1}^T x_{t}^d
$$

### Constrains

$$
x_t = x_t = 0  \text{ for } t \leq -2
$$
$$
x_t = x_t = 1 \text{ for } t \in \{\mathrm{-1}, \mathrm{0}\}
$$
$$
L_{i,t}^{NB} \leq c(x_t^{NB} + x_{t-2}^{SB})
\text{ }, \forall (i \leq 8 \cap \text{ } t >2)
$$
$$
L_{i,t}^{NB} \leq c(x_{t-1}^{NB} + x_{t-3}^{SB})
\text{ }, \forall (i \geq 9 \cap \text{ } t >3)
$$
$$
L_{i,t}^{SB} \leq c(x_{t}^{SB} + x_{t-2}^{NB})
\text{ }, \forall (i \geq 9 \cap \text{ } t >2)
$$
$$
L_{i,t}^{SB} \leq c(x_{t-1}^{SB} + x_{t-3}^{NB})
\text{ }, \forall (i \geq 8 \cap  \text{ } t >3)
$$
$$
4 + \sum_{\tau = t-3}^t 
( x_{\tau}^{NB} + x_{\tau}^{SB} )
\leq 
F, 
\text{ } 
\forall t
$$
$$
x_t^{NB} \leq 9
$$
$$
x_t^{SB} \leq 9
$$
$$
x_t^d \in \mathbb{Z}_{\geq 0}
$$

## 4. Optimisation Approach

The model was implented in Python using Pyomo and solved using GPLK at first and then HiGHS when a more powerful solver was needed. The initial model was implemented and a solution where overcrowding equaled zero was found for the intial capacity of 856. However after adjusting the capacity down to even 428, the model still found a solution where overcrowding was equal to zero. Therefore the model was instead adjusted to minimise the trains dispatches needed for overcrowding to remain at zero. 

## 5. Results

### 5.1 Optimal dispatch schedule and Active fleet

```{r, echo=FALSE}
schedule <- read_csv("Victoria_Monday_Train_opt_results.csv", show_col_types = FALSE)

kable(head(schedule, 15),
      caption = "Snippet of dataframe produced")
```

```{r, echo=FALSE}
vic_mon <- read_csv("Victoria_Monday_Train_opt_results.csv", show_col_types = FALSE)


# Combine directions into one column

vic_mon_merged <- pivot_longer(vic_mon, 
                               cols = c('NB', 'SB'),
                               names_to = 'Direction',
                               values_to = 'Trains_Dispatched')

# Fix times

vic_mon_merged$Start_Time <- vic_mon_merged$Time 

vic_mon_merged <- vic_mon_merged %>% 
  separate(Start_Time, into = c("Start_Time", "End Time"), sep = "-")

vic_mon_merged$Start_Time <- hms::as_hms(
  paste0(substr(vic_mon_merged$Start_Time, 1, 2),":",
    substr(vic_mon_merged$Start_Time, 3, 4),":00"
  )
)



#df for direction
dispatch <- vic_mon_merged[c("Start_Time", "Direction", 
                                  "Trains_Dispatched", "t")]

# df for active
active <- vic_mon_merged[c("Start_Time", "Trains_active", 
                                    "Trains_Dispatched", "t")]


#Plot

ggplot(dispatch, aes(x = t, y = Trains_Dispatched, 
                           group = Direction, 
                     colour = Direction)) + 
  geom_step(linewidth = 1) + 
  geom_point(size = 1) + 
  geom_step(data = active, aes(x = t,
                               y = Trains_active,
                               colour = "Active Trains"),
            inherit.aes = FALSE,
            linewidth = 1
    ) + geom_point(data = active, size = 1, inherit.aes = FALSE,
                   aes(x = t,
                               y = Trains_active,
                               colour = "Active Trains")) + 
  scale_x_continuous(breaks = active$t[seq(1, nrow(active), by = 20)],
    labels = active$Start_Time[seq(1, nrow(active), by = 20)]
  ) +
  theme_minimal() + 
  labs(
    title = "Optimised Train Dispatches",
    x = "15 minute interval",
    y = "Number of Trains") +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "top")


```



## 6. Sensitivity Analysis

### 6.1 Effect of train capacity

```{r, echo=FALSE}
## Bar plot of total trains dispatched vs train capacity

min_total <-  read_csv("Trains_Sensitivity.csv", show_col_types = FALSE)

min_total$Capacity <- as.character(min_total$Capacity)

ggplot(min_total, aes(x = Capacity, y = Minimised_trains,
                      fill = Capacity,
                      colour = Capacity)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Trains dispatched vs Train Capacity",
       x = "Train Capacity", y = "Total Trains Dispatched") +
  theme_minimal()
```


### 6.2 Effect on fleet utilisation

```{r, echo=FALSE}
vic_mon <- read_csv("Victoria_Monday_Train_sensitivity.csv", show_col_types = FALSE)

# Direction to one column

vic_mon <- pivot_longer(vic_mon, 
                               cols = c('NB', 'SB'),
                               names_to = 'Direction',
                               values_to = 'Trains_Dispatched')

# Fix times

vic_mon$Start_Time <- vic_mon$Time 

vic_mon <- vic_mon %>% 
  separate(Start_Time, into = c("Start_Time", "End Time"), sep = "-")

vic_mon$Start_Time <- paste0(
  substr(vic_mon$Start_Time, 1, 2),
  ":",
  substr(vic_mon$Start_Time, 3, 4)
)

vic_mon$Capacity <- as.character(vic_mon$Capacity)

# Plot trains active for each interval, split by capacity

ggplot(vic_mon, aes(x = t, 
                    y = Trains_active,
                    group = Capacity, 
                    colour = Capacity)) + 
  geom_step(linewidth = 1) + 
  geom_point(size = 1) + 
  labs(
    title = "Minimised Active Trains for each Capacity and Interval",
    x = "Time",
    y = "Number of Trains Active"
  ) +
  scale_x_continuous(breaks = vic_mon$t[seq(1, nrow(vic_mon), by = 80)],
                     labels = vic_mon$Start_Time[seq(1, nrow(vic_mon), by = 80)]) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold")
  )
```
```{r, echo=FALSE}
# Dispatch schedule for selected capacities

#df for direction
dispatch <- vic_mon[c("Start_Time", "Direction", 
                             "Trains_Dispatched", "t", "Capacity")]

# df for active
active <- vic_mon[c("Start_Time", "Trains_active", 
                           "Trains_Dispatched", "t", "Capacity")]

# create function for plot

plot_capacity <- function(capacity) {
  
  plot_data1 <- dispatch %>%
    filter(Capacity == capacity)
  
  plot_data2 <- active %>%
    filter(Capacity == capacity)
  
  ggplot(plot_data1, aes(x = t, y = Trains_Dispatched, 
                       group = Direction, 
                       colour = Direction)) + 
    geom_step(linewidth = 0.5) + 
    geom_point(size = 0.5) + 
    geom_step(data = plot_data2, aes(x = t,
                                 y = Trains_active,
                                 colour = "Active Trains"),
              inherit.aes = FALSE,
              linewidth = 0.5
    ) + geom_point(data = plot_data2, size = 0.5, inherit.aes = FALSE,
                   aes(x = t,
                       y = Trains_active,
                       colour = "Active Trains")) + 
    scale_x_continuous(breaks = plot_data2$t[seq(1, nrow(plot_data2), by = 60)],
                       labels = plot_data2$Start_Time[seq(1, nrow(plot_data2), by = 60)]
    ) +
    theme_minimal() + 
    labs(
      title = paste("Optimised Train Dispatches
                    at capacity", capacity),
      x = "15 minute interval",
      y = "Number of Trains") +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "top")
}

p1 <- plot_capacity(856)
p2 <- plot_capacity(749)
p3 <- plot_capacity(642)
p4 <- plot_capacity(428)

combined_plot <- p1 + p2 + p3 + p4

combined_plot
```


## 7. Limitations

## 8. Conclusion
